---
title: "L2.4: Transformations of MLEs"
subtitle: "BIOS 6612"
author: "Julia Wrobel"
date: February 1, 2021
header-includes:
   - \usepackage{bm}
output:
  powerpoint_presentation:
    reference_doc: ../cu_style.pptx
---

```{r, echo= FALSE, include = FALSE}
library(tidyverse)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)

theme_set(theme_bw() + theme(legend.position = "bottom"))
```

## Overview

Today, we cover:

* Invariance property of MLEs
* Inference on transformations of MLEs
* The Delta Method

<br>

Optional readings:

* Dobson & Barnett: Chapter 5.3
* Agresti: 3.1.5-3.1.7
* Casella and Berger: pages 319-322


## Invariance Property of MLEs

Allows us to find the MLE of transformations of an MLE

<br>

* If $\hat{\theta}$ is the MLE of $\theta$, then for any function $\tau(\hat{\theta})$ is the MLE of $\tau(\theta)$!



::: notes
- Show an example on the next slide
- I will not go over the proof, but you can find it in Casella and Berger
::: 

## Invariance Property of MLEs

Suppose $Y_1, Y_2, \ldots, Y_n$ is a sample of independent Normal($\mu$, $\sigma^2$) random variables with $E(Y_i) = \mu$.

* Sample mean is $\hat{\mu} = \bar{Y} = \frac1n \sum_{i = 1}^n Y_i$ is the MLE of $\mu$

<br>

What is the MLE of $1/\mu$? Using invariance property of MLEs,

* $1/\hat{\mu} = 1/\bar{Y}$ is the MLE of $1/\mu$


## Variance of MLE transformations

* What if we want a confidence interval around our estimate $1/\hat{\mu}$?

<br>

* How do we calculate $Var[\tau(\theta)]$?
  * Use the Delta Method

##  Delta Method

* Assume $\sqrt{n}(\hat{\theta}_n - \theta_0) \to^d N(0, \sigma^2)$
  * Central Limit Theorem, Asymptotic normality of MLEs
  
<br>  
  
* For a given function $\tau$ and specific value of $\theta_0$, suppose that $\tau'(\theta_0)$ exists and is not 0
* Then 

$$\sqrt{n}(\tau(\hat{\theta}_n) - \tau(\theta_0)) \to^d N(0, [\tau'(\theta_0)]^2\sigma^2)$$


* So asymptotic variance $Var[\tau(\theta)] \approx [\tau'(\theta_0)]^2\sigma^2)$

::: notes
- Delta method allows you to find asymptotic variance of functions of estimators
- Approx equal and not directly equal because this is asymptotic variance, only true in the limit
::: 


## Delta method: example


* Take our earlier estimate of $1/\mu$, $1/\bar{Y}$
* From central limit theorem, we know that 

$$\sqrt{n}(\bar{Y}- \mu) \to^d N(0, \sigma^2)$$

* $\sigma^2 = Var(Y_1)$
* Our function $\tau$ is $\tau(\mu) = 1/\mu$
  * Derivative is $\tau'(\mu) = -\frac{1}{\mu^2}$
* Delta method tells us that 

$$\sqrt{n}(1/\bar{Y}- 1/\mu) \to^d N\left[0, \left(-\frac{1}{\mu^2}\right)^2\sigma^2\right]$$

* So variance of $1/\bar{Y}$ is $\frac{\sigma^2}{\mu^4}$
  
::: notes
- Delta method allows you to find asymptotic variance of functions of estimators
:::  
  
## Delta method: example

* We can check this answer using simulations!! Try $Y_1 \sim N(0.5)$

```{r, echo = TRUE}
trials = 100000 
tau_theta_hat = var_tau_theta_hat = rep(NA, length.out = trials) # define empty vector
n = 100

set.seed(20341)

for(trial in 1:trials) { 
  Y = rnorm(n, mean = 0.5)
  tau_theta_hat[trial] = 1/mean(Y)    # calculate 1/pi_hat
  var_tau_theta_hat[trial] = (var(Y)/n)/mean(Y)^4 # calculate delta method variance
}
```

## Delta method: example

```{r, echo = TRUE}
mean(tau_theta_hat)
# should be equal to 
1/.5

var(tau_theta_hat)
# should be equal to 
mean(var_tau_theta_hat) 
```

* We get reasonably close to the calculated theoretical values when $n = 100$
* Look at the distribution of these values in the next slide
  * Should be $\tau(\theta) \sim N(2, 0.4)$

## Histogram of t(theta)


:::::::::::::: {.columns}
::: {.column}
```{r, echo=FALSE, fig.height = 3}
tibble(tau_theta_hat = tau_theta_hat) %>%
  ggplot(aes(tau_theta_hat)) +
  geom_histogram(aes(y = ..density..), fill = "lightblue") +
  stat_function(fun = dnorm, args = list(mean = 1/0.5, sd = sqrt((1/n)/.5^4)))
```
:::
::: {.column}
* The histogram doesn’t seem to adhere too closely to the normal density curve, which could suggest our sample size of $n = 100$ is too small for the asymptotic results to hold very exactly

<br>

* Let’s simulate again with $n = 1000$
:::
::::::::::::::




## Histogram of t(theta)

:::::::::::::: {.columns}
::: {.column}

```{r, echo = FALSE, fig.height = 3}
trials = 100000 
tau_theta_hat = var_tau_theta_hat = rep(NA, length.out = trials) # define empty vector
n = 1000

set.seed(20341)

for(trial in 1:trials) { 
  Y = rnorm(n, mean = 0.5)
  tau_theta_hat[trial] = 1/mean(Y)    # calculate 1/pi_hat
  var_tau_theta_hat[trial] = (var(Y)/n)/mean(Y)^4 # calculate delta method variance
}

tibble(tau_theta_hat = tau_theta_hat) %>%
  ggplot(aes(tau_theta_hat)) +
  geom_histogram(aes(y = ..density..), fill = "lightblue") +
  stat_function(fun = dnorm, args = list(mean = 1/0.5, sd = sqrt((1/n)/.5^4)))

```

:::
::: {.column}
* This is better: sometimes it can take very large samples before asymptotic results “kick in”

:::
::::::::::::::



