---
title: "Lecture 3.3: Logistic Regression Implementation"
subtitle: "BIOS 6612"
author: "Julia Wrobel"
date: February 8, 2021
header-includes:
  - \usepackage{bm}
output:
  powerpoint_presentation:
    reference_doc: ../cu_style.pptx
---
  
```{r, echo= FALSE, include = FALSE}
## generalizing the normal linear model to


library(tidyverse)
library(dobson) 
#  data from Dobson and Barnett GLM book

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

theme_set(theme_bw() + theme(legend.position = "bottom"))
```


## Overview

Today, we cover:

* Logistic regression assumptions
* Grouped (group-level) and ungrouped (subject-level) data  


<br>
  
Readings:
  
* Agresti: 5 up to 5.23, 5.4



## Multiple Logistic Regression

Logistic regression with multiple covariates.

* $\pi_i = P(Y_i  = 1)$ and $\boldsymbol{x}_i\boldsymbol{\beta} = \beta_0 + \beta_1X_{1i} +\ldots + \beta_pX_{pi}$

<br>

Model takes the form

$$\log\left(\frac{\pi_i}{1-\pi_i}\right) = \boldsymbol{x}_i\boldsymbol{\beta} 

\implies 

\pi = \frac{e^{\boldsymbol{x}_i\boldsymbol{\beta}}}{1 + e^{\boldsymbol{x}_i\boldsymbol{\beta}}}$$

<br>

Interpretation is

* A unit increase in $X_j$ multiplies the odds by $e^{\beta_j}$
* A unit increase in $X_j$ multiplies the log-odds by $\beta_j$

## Logistic Regression Assumptions


* Observations are independent

<br>

* There is a linear relationship between the covariates $X$ and the log odds of $Y=1$

<br>

* For subject $i$ $Y|X_i$ is Bernoulli with $P(success) = P(Y_i+1|X_i= x_i) = \pi(x_i)$ and variance $\pi(x_i)[1-\pi(x_i)]$
  * Note: $\pi(x_i)$ is a probability that is a function of the covariates
  * Note: The variance is *also* a function of the covariates (not homoscedastic)!




## Group vs. Subject Level Data


:::::::::::::: {.columns}
::: {.column}
Group-level data

* Outcome: Binary (1 = has disease, 0 = doesn't have disease)
* Predictors: All predictors are discrete (no continuous predictors)
* Notation: $Y_i$ refers to group $i$'s # of successes

$$Y_i \sim Binomial(n_i, \pi_i)$$
:::
::: {.column}
Subject-level data

* Outcome: Binary (1 = has disease, 0 = doesn't have disease)
* Predictors: At least one continuous predictor
* Notation: $Y_i$ refers to subject $i$'s success status

$$Y_i \sim Binomial(1, \pi_i) = Bernoulli(\pi_i)$$
:::
::::::::::::::

## Relationship between Binomial and Bernoulli distributions

The Binomial($n, \pi$) distribution models the number of times $Y = 1$ (usually termed success) in an experiment that occurs $n$ times, where $\pi = P(Y = 1)$

<br>

* A Binomial distribution is the sum of i.i.d. Bernoulli random variables

<br>

* A Binomial random variable with $n = 1$ (one trial) is equivalent to a Bernoulli random variable


## Group vs. Subject Level data implications

How we treat and test for these will not always be exactly the same

<br>

* Depending on which type of data layout you have, `R` code for fitting a logistic model will change.

<br>

* Some analyses of how well the model fits the data (deviance) will depend on the type of data you have


## Group Level data 

All data can be summarized in a table. Assume a binary out come and two binary predictors, gender $\in$ {male, female} and age $\in$ {young, old}

<br>

* Each set of covariates is called a "covariate pattern"
  * gender = "female" and age = "young"
  * gender = "male" and age = "young" 
  * gender = "female" and age = "old"
  * gender = "male" and age = "old" 
  
<br>

If there are $m$ distinct covariate patterns, for each pattern we record the number of individuals having that pattern ($n_j$) and the number of success $s_j$ for $j = 1,\ldots,m$.


## Likelihood for Group Level Data

Assume $m$ different covariate patterns, and for pattern $j$ $s_j \sim Binomial(n_j, \pi_j)$


$$f(s_j; \pi_j) = {n_j\choose s_j} \pi_j^{s_j}(1-\pi_j)^{n_j - s_j}$$

<br>

$$L(\pi_1,\pi_2,\ldots, \pi_m) = \prod_{j = 1}^m {n_j\choose s_j} \pi_j^{s_j}(1-\pi_j)^{n_j - s_j} \propto \prod_{j = 1}^m \pi_j^{s_j}(1-\pi_j)^{n_j - s_j}$$

::: notes
- Likelihood for subject-level data is the same one we derived last lecture
- Likelihood is a function of each (independent) covariate pattern
- When you maximize, get same estimators as you would for individual level data
::: 

## Group Level Example: Budworm data

The data come from an experiment on the toxicity of doses of a poison (trans-cypermethrin) to the tobacco budworm (a moth that eats tobacco and other crops).

* Batches of 20 moths of each sex were exposed for three days to the poison 
* Each batch received different dose of poison
  * 6 doses (1,2,4,8,16,32 mg)
* The number in each batch that died was recorded ("success" = death)


Data is grouped

* 6 doses, 2 sexes
* $m = 12$ covariate patterns
* $n_j = 20$ moths in each covariate pattern

## Group Level Example: Budworm data

* sex: 1 = male, 0 = female
* dose: mg of poison received
* s: # of budworms killed
* n: total # of budworms


```{r, echo = TRUE}
#budworm <- read.table("budworm_data.txt",header=TRUE) # alternative way to load the data
budworm = read_table2("budworm_data.txt", col_names = TRUE)
budworm %>% print(n = 3)

nrow(budworm) # number of rows = number of groups
```

## Budworm data EDA

```{r}
budworm %>%
  mutate(prop_dead = s/20,
         sex_factor = factor(sex, levels = 0:1, labels = c("female", "male"))) %>%
  ggplot(aes(dose, prop_dead)) +
  geom_point() +
  labs(x = "Poison dose in mg", y = "proportion of budworms killed") +
  facet_wrap(~sex_factor)
  
```


## Budworm data model

```{r, echo = TRUE}
budworm_mod = glm(cbind(s, n-s) ~ sex + as.factor(dose), 
                    family=binomial, data = budworm)
broom::tidy(budworm_mod)
```


## Subject Level Example: Birthweight data

A study was conducted to identify risk factors for low infant birth weight using data from 189 live births at Bay State Medical Center in Massachusetts. The data file *lowbwt.csv* has the following variables.

* id: id number of infant
* low: the outcome (1 = birthweight low < 2500 grams) vs (0 = not low > = 2500 grams)
* smoke: smoking during pregnancy, 1 = yes; 0 = no
* age: mother's age in years
* visits: number of physician visits during 1st trimester
  * 0: none
  * 1: one
  * 2: two or more

Age is continuous so this data cannot be summarized at the group level. Researchers are interested in how mother's age and smoking status influences probability of low birthweight.

## Birthweight data 

```{r, echo = TRUE}
low_birthweight = read_csv("lowbwt.csv")

low_birthweight %>% print(n = 3)
nrow(low_birthweight) # number of rows = number of subjects
```



## Birthweight data 

```{r, echo = TRUE}
birthweight_mod = glm(low ~ smoke + age, family = binomial, data = low_birthweight)
broom::tidy(birthweight_mod)
```



