---
title: "BIOS 6612: A very brief intro to the tidyverse"
output:
  html_document: 
    toc: true
    toc_float: true
---

# Overview

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

library(tidyverse)
theme_set(theme_bw() + 
          theme(legend.position = "bottom"))
```

In this course, I will do examples in `R` using `tidyverse` syntax. The `tidyverse` is a suite of `R` packages developed by RStudio that make it easier to load, clean, manipulate, and visualize data. You can load multiple `tidyverse` packages simulatneously using the following line of code.

```{r, message = FALSE}
library(broom)
library(tidyverse)
library(janitor)
```


# Course stuff

* My lecture slides are written in `Rmarkdown` and **knitted** to Powerpoint
* Some data examples come from the `dobson` package which links to examples from [this book](http://www.ru.ac.bd/wp-content/uploads/sites/25/2019/03/202_06_Dobson_An-Introduction-to-Generalized-Linear-Models-2008.pdf).
* Make sure to fill out the survey on Canvas to pick TA office hours!
* [Schedule a one-on-one meeting with me as well!](https://calendly.com/julia-wrobel/15min)


## TAs

* Ying Jin
  * 2nd year Biostats PhD Student
  * RA in CIDA
  * Got a pet guinea pig during COVID!
  
* Xin (Randy) Jin
  * 2nd year Biostats PhD Student
  * Interested in mixed models and Bayesian analysis

# The `tidyverse` versus base R

"base R" is the version of `R` you get when you install `R` but don't install any other libraries. 

* Most of the data stuff you can do in the `tidyverse` you can also do in base R
* If you want to use base R instead, that's fine and you won't be penalized on homework assignments

Reasons I'll be using `tidyverse` syntax

* I like it
* It's easier to read
* It's good to get exposure to different styles of coding

I won't be teaching `tidyverse` concepts in detail because it's beyond the scope of our class, but I am always happy to answer questions.


## Installing the `tidyverse`

```{r, eval = FALSE}
# install.packages("tidyverse")
library(tidyverse)
```

This installs and loads a suite of packages, including

* `dplyr`: data manipulation
* `ggplot2`: data visualization
* `tidyr`: data cleaning
* `readr`: reading in data
* `purrr`: functional programming
* `tibble`: tidy version of a `data.frame`
* `stringr`: working with strings/character valued data
* `forcats`: for categorical variables, deals with factors

Cheatsheets are available online for each of these packages, I added the `dplyr` and `ggplot2` to the resources page on Canvas.

## The pipe operator `%>%`

The pipe operator (part of the `tidyverse`) let's you chain together functions in `R`. Say you want to do a series of operations in a row. See two ways below:


```{r}
x <- c(0.109, 0.359, 0.63, 
       0.996, 0.515, 0.142,
       0.017, 0.829, 0.907)

# (1) can do this
round(exp(diff(log(x))), 1)

# (2) or this
logx = log(x)
diff_lx = diff(logx)
exp_df_lx = exp(diff_lx)
round_result = round(exp_df_lx, 1)
```

The first way is hard to read, and the second way saves intermediate variables (`logx`, `diff_lx`, ... ) you don't need.

Alternatively, you could use the pipe operator:

```{r}
x %>% log() %>%
    diff() %>%
    exp() %>%
    round(1)
```


## tibbles

A _tibble_, or `tbl_df`, is basically just a tidyverse version of the `data.frame`. They have some nice properties that make them behave a little bit differently from a data.frame:

* Doesn't automatically convert strings to factors
* Doesn't use rownames
* Only prints first 10 rows and number of columns you can fit on one screen:

```{r}
mtcars
tibble(mtcars)
```

You can think about it as a dataframe with a few different properties. Sometimes, when you do tidyverse operations on your dataframe it gets converted to a tibble anyway.


## `readr`

I'm going to use data from [Denver Open Data](https://www.denvergov.org/opendata) to highlight utility of the tidyverse. This website provides datasets on all sorts of weird Denver-municipal stuff, including

* Crime related to marijuana use
* Geolocation of swimming pools in Denver County
* City budgets
* Location and checkout history for Denver libraries

I'm going to use the libraries data set. You can download a csv of it [here](https://www.denvergov.org/opendata/dataset/city-and-county-of-denver-libraries).

You can read in the csv file directly from the websit using the `read_csv` function from the `readr` package. 

```{r}
libraries = read_csv("https://www.denvergov.org/media/gis/DataCatalog/libraries/csv/libraries.csv")
libraries
```

This produces a `tibble`.  Alternatively, you could read in the data using `read.csv`.

I like my variable names to be in a specific format- all lowercase with no spaces in the varible names. Luckily, there is a function in the `janitor` package that will do this for you.

```{r}
libraries = libraries %>% janitor::clean_names()
libraries
```


I do this basically every time I load a dataset into `R`.


## `dplyr` verbs

`dplyr` is my favorite set of tools for data manipulation. It is made up of **verbs** (which are `R` functions) that each solve a different data manipulation task. I'm just showing you a common subset of the verbs I use here, there are many more!

### `select`

Here are the variables available in our libraries dataset.

```{r}
names(libraries)
```

`select` lets you choose a subset of variables. Let's say I only care about a few of the variables.

```{r}
libraries_subset = libraries %>% 
  ## to tidy the dataset
  select(library_name, city, total_bldg_sqft) 
libraries_subset
```

FYI - you don't have to use the `%>%` operate with these `select` or any of the other variables below. I just like the flow of it. Alternatively, code below does the same thing as above:

```{r}
libraries_subset <-  
  select(libraries, 
         library_name, 
         city, 
         total_bldg_sqft)
```


You can also use `select` to remove variables

```{r}
libraries %>% 
  select(-status)
# libraries = libraries %>% select(-status) to save
```

Or to reorder variables

```{r}
libraries %>%
  select(library_name, 
         starts_with("address"), 
         door_count, 
         everything())
```


### `filter`

`filter` let's you choose a subset of rows, based on whether criteria that you supply are TRUE.

```{r}
libraries %>%
  filter(circulation > 600000)
```

You can combine different conditions

```{r}
libraries %>%
  filter(city == "Denver", 
         branch_code %in% c("BDY", "UNH", 
                            "CRK", "FIE"), 
         door_count > 80000)
```


### `mutate`

`mutate` allows you to create new variables and add them to you dataset

```{r}
libraries_new <-
  libraries %>%
  mutate(high_circulation = 
           ifelse(circulation > median(circulation), TRUE, FALSE),
         favorite_food = "pizza",
         double_door_count = door_count * 2) %>%
  select(high_circulation, 
         favorite_food, 
         double_door_count, 
         everything())

libraries_new
```

### `summarize` and `group_by`

`summarize` produces a summary of variables in your dataset. `group_by` allows you to do operations by group. I often use it in conjunction with `summarize`.

```{r}
libraries %>%
  summarize(median = median(circulation), 
            mean = mean(circulation), 
            variance = var(circulation))
```


grouping by zip, calculating number of branches in that zip code. The `arrange` verb sorts the rows.

```{r}
libraries %>%
  group_by(zip) %>%
  summarize(n_branches = n()) %>%
  arrange(-n_branches)
```


Combining some of the verbs above and producing a prettier table. Make sure to `ungroup()`!

```{r}
libraries_new %>%
  mutate(high_circulation = ifelse(circulation > median(circulation), 
                                   TRUE, FALSE)) %>%
  group_by(high_circulation) %>%
  summarize(mean_count = mean(door_count), 
            mean_sqft = mean(total_bldg_sqft)) %>%
  ungroup() %>%
  knitr::kable()
  
```


`dplyr` verbs are intended to be used together!

## `ggplot2`

Many of you may be familiar with `ggplot2`, but I'll show a couple examples.

* Histogram

```{r}
libraries %>%
  ggplot(aes(x = circulation)) +
  geom_histogram() +
  ggtitle("Circulation of books in Denver libraries")
```

FYI, if you're not using the pipe operator:

```{r, eval = FALSE}
ggplot(libraries, 
       aes(x = circulation)) +
  geom_histogram() +
  ggtitle("Circulation of books in Denver libraries")
```

Scatterplot- the Central branch is by far the biggest.

```{r}
libraries %>%
  ggplot(aes(x = circulation, 
             y = total_bldg_sqft)) +
  geom_point(color = "orange") +
  geom_text(aes(label = branch_code))
```

Scatterplot with  a regression line- I took the Central branch out of the data as well.

```{r}
libraries %>%
  filter(branch_code != "CEN") %>%
  ggplot(aes(x = circulation, 
             y = total_bldg_sqft)) +
  geom_point(color = "orange") +
  geom_smooth(method = "lm", se = FALSE)
```


Facet_wrap

```{r}
libraries_new %>%
  filter(branch_code != "CEN") %>%
  ggplot(aes(x = door_count, 
             y = total_bldg_sqft)) +
  geom_point(color = "orange") +
  geom_smooth(method = "lm", se = FALSE)  +
  facet_wrap(~high_circulation)
```



# Wrap up

* That's all for today
* I'm happy to talk about the rest of the packages in the tidyverse in subsequent labs- just let me know
* Links to cheat sheets and tutorials are on Canvas
